---
title: "Database_format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Database_format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

```


**Table 1. database format. "field" is variable name used in FishStomachs; "types" is the type of input(c=character, i=intger, d=double, l=logical);	"mandatory" is the field required as minimum (T=TRUE, F=FALSE);	"PRED" is the variable part of predator information;	"PREY" is the informtion part of prey information; "units" is the unit or values applied.**
```{r, maketable1, results = "asis", echo=FALSE,message=FALSE}

a<-readr::read_csv(file.path("..","inst","stomach_format.csv"),col_types = readr::cols()) 
a<-  dplyr::mutate_if(a,is.logical,as.character()) 
a<-  dplyr::mutate(a,mandatory=dplyr::if_else(mandatory=='FALSE','F','T'),
                   PRED=dplyr::if_else(PRED=='FALSE','F','T'),PREY=dplyr::if_else(PREY=='FALSE','F','T'))
a<-  dplyr::mutate(a,units=dplyr::if_else(is.na(units),' ',units))
knitr::kable(a[,4:10])

```


**Table 2. Alias names accectped as default. Can be changed by using a local version of the the database fortmat  (see xxxxc).**
```{r, maketable2, results = "asis", echo=FALSE,message=FALSE}
a<-dplyr::mutate(a,alias_1=dplyr::if_else(is.na(alias_1),' ',alias_1),alias_2=dplyr::if_else(is.na(alias_2),' ',alias_2),alias_3=dplyr::if_else(is.na(alias_3),' ',alias_3))
knitr::kable(a[,1:4]) 

```

