---
title: "FishStomachs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FishStomachs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Compilation of stomach content data for estimation of population diet and food ration. ##

Compilation of data from individual stomachs to population level is not
trivial and the method will depend highly on the question asked. Stomach
data are often collected to get information on the average diet or food
ration of a given species within a given area to inform multispecies
assessment models (e.g. SMS or Gadget). Such models require data on diet
and biomass eaten (food ration) by predator species and size
classes. The average "population" diet or food ration should basically
be calculated from a stratified mean of the individual stomach content
samples, weighted by strata adensity of the predator and the area of the strata.
This seems simple, but incomplete and patchy sampling makes it often
necessary to use a series of *ad hoc* solutions.

In general a series of data compilation methods could be applied:

1.  Read and check data from agreed exchange format;
2.  Bias correct to take into account variable evacuation rate;
3.  Assign size classes for predators and preys;
4.  Bias correct to take into account regurgitated stomachs within sample units;
5.  Aggregate stomachs contents within sample_id and size classes.
6.  Allocate unidentified or partly identified prey items;
7.  Calculate population diet and food ration from a weighted average. 

The sequence of these steps, of which some of the steps might be repeated, 
will depend on the individual sampling design and the quality of the analysed data.

The FishStomachs package defines data structures suitable for stomach data, and provides the necessary methods to compile observed stomach data into population diet and biomass eaten, used for multispecies models. The methods applied for a set of observations are stored within the data output to document the compilation steps taken. 

With my experience from compiling stomach data for the North Sea for use
in the MSVPA/SMS models, this document describes the methods above and
how these can be applied.

### 1) Data exchange format ###

Data must be recorded, stored and exchanged on an agreed and documented
format suitable for the purpose of sampling. 
This R-package uses a user defined format (see Table 1)  with the possibility to rename variable names
such that it is possible to include existing
data bases (e.g. ICES stomach contents data base). The new format includes a number of mandatory (system defined) and optional (user defined) variable names, 
which uses "aliases" (See Table 2)  as a link to "internal" variable names used for processing by FishStomachs. 


**Table 1. Database format. "field" is variable name used in FishStomachs; "types" is the type of input(c=character, i=integer, d=double, l=logical);	"mandatory" is the field required as minimum (T=TRUE, F=FALSE);	"PRED" is the variable part of predator information; 
"PREY" is the information part of prey information; "units" is the unit or values applied.
The data base format is specified in the file stomach_format.csv.**

```{r, maketable1, results = "asis", echo=FALSE,message=FALSE}
a<-readr::read_csv(file.path(system.file( package = "FishStomachs"),"stomach_format.csv"),col_types = readr::cols())
a<-  dplyr::mutate_if(a,is.logical,as.character()) 
a<-  dplyr::mutate(a,mandatory=dplyr::if_else(mandatory=='FALSE','F','T'),
                   PRED=dplyr::if_else(PRED=='FALSE','F','T'),PREY=dplyr::if_else(PREY=='FALSE','F','T'))
a<-  dplyr::mutate(a,units=dplyr::if_else(is.na(units),' ',units))
knitr::kable(a[,4:10])

```
  
     
  
**Table 2. Alias names accepted as default. Aliases can be changed and added  by using a local version of the the database format. The data base format and aliases are specified in the file stomach_format.csv.**
```{r, maketable2, results = "asis", echo=FALSE,message=FALSE}
a<-dplyr::mutate(a,alias_1=dplyr::if_else(is.na(alias_1),' ',alias_1),alias_2=dplyr::if_else(is.na(alias_2),' ',alias_2),alias_3=dplyr::if_else(is.na(alias_3),' ',alias_3))
knitr::kable(a[,1:4]) 

```
   


The presently used ICES
data exchange format for stomachs (see Appendix 1 in this document) is based on the "Year
of the stomach" sampling for North Sea stocks in the period 1981-1991
(see Appendix 2). The new version includes some extensions to better
include stomachs sampled outside the "Year of the stomach", e.g. from
national sampling programs with less standardized recording. Another
change is that the new exchange format uses the WoRMS AphiaID
(<http://www.marinespecies.org/index.php>) species taxonomic ode, where
the old format used the NODC codes
(<https://www.nodc.noaa.gov/General/CDR-detdesc/taxonomic-v8.html>).


In this document, the examples and variable names used are based on a
data set on the ICES exchange format (Appendix 1).

### 2) Bias correction to take into account variable evacuation rate ###

It is often assumed that diet (eaten food) is the same as the observed stomach
content. This is however not the case as the stomach evacuation rate of
an individual stomach depends on e.g. the average meal size, average
energy density of the meal, prey armament and temperature. Large energy
rich meals will for example have lower evacuation rate than small energy
poor meals. Therefore, large energy rich meals will be present in a
stomach a longer time and thereby detectable for a longer time, than
stomach contents from small energy poor meals.

In the two sub-sections below it is described how to bias correct the
effect of variable stomach evacuation rate and thereby how to bias
correct the estimated average diet (or consumption rate) estimated from
the observed stomach contents data. Ideally, this bias correction should
take place at the level of sampling, i.e. by the individual stomach or
by the pool of stomachs by sample site.

#### Estimation of food ration from stomach contents data ####

Food rations (evacuation rate of stomach contents) can be estimated from
the observed stomach contents using the methods suggested by Andersen
and Beyer (2005 a, b). This model takes into account the differences in
evacuation rates between prey types due to their energy density and
their resistance to digestion (armament).

Ration (R) (per hour) of prey group (i) for pooled stomachs is
calculated from:


$$R = \sum_{i}^{}{(\rho*M_{i}*b_{i}*e^{\delta T}*L^{\lambda}}*E^{-\xi})*K*\left(\frac{N_{A}}{N_{F}} \right)^{\alpha - 1}*S^{\alpha}$$

M= armament of individual prey (group) i

b=proportion of prey (group) i

T= temperature (^o^C)

L= length (cm) of the predator

E= average energy density (kJ/g wet weight) of the stomach (or of the
pooled stomachs)

N= Number of stomachs in the sample, total (A) and with food (F)

S = average stomach contents in grams


$\rho$, $\delta$, $\lambda$, $\xi$ and K are parameters to the model (Table 3)

Values for armament (M) and average energy density of individual prey
(groups) as used in the compilation of the North Sea stomachs are presented
in Table 5.

When data exist by the individual stomachs, the ration eaten per
(feeding) fish can be calculated as:

$$R = \sum_{i}^{}{\rho * M_i* b_i* e^{\delta T} * L^{\lambda} * E^{-\xi} * S^{\alpha}}$$

**Table 3. Parameter values of the generic cylinder model of gastric
evacuation used for the compilation of stomachs from predators in the
North Sea.**

  Predator   $\rho$    $\lambda$ $\delta$  $\xi$  $\alpha$   K
  ---------- --------- --------  -------- ------- -------- ------
  Cod        0.00224   1.30      0.083    -0.85   0.5      0.85
  Haddock    0.00191   1.30      0.083    -0.85   0.5      0.85
  Saithe     0.00171   1.35      0.081    -0.85   0.5      0.85
  Whiting    0.00171   1.35      0.081    -0.85   0.5      0.85
  Mackerel   0.00174   1.30      0.080    -0.85   0.5      0.85

#### Estimation of diet from stomach contents data ####

The method for bias correction in the calculation of food ration as
outlined above can also be applied to calculate the ration of the
individual prey item and thereby the average diet. For example, for
individual stomachs the Ration (R) of prey item *i* becomes:

$$R_{i} = \rho\ M_i *b_i * e^{\delta T} * L^\lambda * E^{-\xi}$$

Based on the set of $R_{i}$ for the individual predators from e.g. a
sampling site (trawl haul) the average Ration of the of the individual
prey item can be calculated by taking the number of feeding, empty and
regurgitated stomachs into account. This average Ration by prey group is
in this case expressed as g wet weight per hour (with $\rho$ included in
the equation) and can easily be aggregated with average rations of preys
from other strata. These rations can finally be transformed into diet
data (weight proportions by prey item).

**Table 4. Average energy density and armament of prey groups as applied
in the compilation of stomach for use in the North Sea SMS.**

```{r, maketable4, results = "asis", echo=FALSE,message=FALSE}
a<-readr::read_csv(file.path(system.file( package = "FishStomachs"),"extdata","energy_density_and_armament.csv"),col_types = readr::cols())

knitr::kable(a,col.names=c("Species","Quarter","Lower length (mm)","Upper length (mm)","Energy density (kJ/g wet weight)", "Armament"))

```


### 3) Assign predator and prey size classes. ###

Stomach data can be sampled, analysed and recorded by the individual
fish or by size group ("pooled stomach data") for the individual sample
unit (sample_id), e.g. trawl haul. To save time during sampling and analysis of the
stomach contents, the "Year of the stomach" data were sampled and
analysed by pre-determined size classes (Table 4). Pooled stomachs do
not provide information by the individual fish and on the variation in
stomach contents, which is preferable in e.g. estimation of daily food
ration and diet. Grouping of stomachs by size classes is however often
necessary in the later compilation of data, as data are to scarce to be
compiled by e.g. each cm group separately.

In general, I will suggest to use size classes, e.g. the size
classes used in 1991 by "Year of the stomach" sampling (Table 3). That
means that the variables in the data exchange format.

-   pred_ll (predator lower length bound)

-   pred_lu (predator upper length bound)

-   prey_ll (prey lower length bound)

-   prey_lu (prey upper length bound)

are set from the length information on the individual predators
and preys, if not already included in the data and defined by the sampling design.


**Table 5. Size classes (for items larger than 5 cm) used for sampling and
compilation of stomach data in the North Sea "Year of the stomach" 1991
sampling.**

  Length class, 1991   Length (cm)
  -------------------- -------------
  50                   5-5.9
  60                   6-6.9
  70                   7-7.9
  80                   8-9.9
  100                  10-11.9
  120                  12-14.9
  150                  15-19.9
  200                  20-24.9
  250                  25-29.9
  300                  30-34.9
  350                  35-39.9
  400                  40-49.9
  500                  50-59.9
  600                  60-69.9
  700                  70-79.9
  800                  80-99.9
  1000                 100-119.9
  1200                 \>120

### 4) Bias correction: Taking number of empty and regurgitated stomachs into account ###

Ideally each stomach should be classified as: with food, with food but
regurgitated, with skeletal remains only or empty. This information is
included in the presently used exchange format.

 
Database variable        Comments
---------------------    -----------------------------------------------------------------
n_food                    Stomachs with (recently ingested) food
n_regur                   Stomach with food, but evidence of regurgitation of parts or all the stomach contents.
n_skel                    Stomachs with practically indigestible remains and the fish is not considered feeding recently
n_empty                   No stomach contents
 
Stomach contents from regurgitated (or everted) stomachs should not be
included in the estimation of diet as the proportion regurgitated is
unknown. The number of fish with regurgitated stomachs should however be
recorded ("n_regur"). With the assumption that the
regurgitated stomachs sampled within a "sample_id" have had the same
stomach contents as the feeding (non-regurgitated) fish each observed
prey item weight and number should be corrected by a factor to calculate
the mean stomach contents of a predator within "sample_id".

$factor =  \frac{(N_{F} + N_{R})}{N_{F}*(N_{F} + N_{SR} + N_{R} + N_{E})}$

Where

$N_{F}$ = Number of feeding fish ("n_food")

$N_{R}$ = Number of feeding fish, but regurgitated
("n_regur")

$N_{\text{SR}}$ = Number of non-feeding fish, stomachs containing only
skeletal remains ("n_skel")

$N_{E}$ = Number of non-feeding fish with empty stomachs
("n_empty")

In practice it is difficult to detect fish with regurgitated stomachs,
and later data analysis may often show that there is a clear difference
in the proportions of regurgitated stomachs between vessels surveying in
the same area at the same time.

Another issue is how to distinguish "n_skel" from
"n_food". In the original compilation of the 1991 "Year of
the Stomach" (ICES, 1997), $N_{SR}$ was added to the numerator in
the calculation of the factor above. However I think
"n_skel" should be considered as stomachs from not
(recently) feeding fish and should be treated as empty stomachs, i.e.
that the stomach contents should be disregarded in the compilation of
data for feeding fish.

### 5) Aggregate stomach contents within sample_id and size classes ###


### 6) Allocate unidentified and partly identified prey items ###

Parts of the stomach contents are often not fully identified to species
(group) or size class. With the assumption that the unidentified items
follow the same species and size distribution as the identified prey
items, it is possible to allocate the unidentified items proportionally
to the identified species and size class. However, with the same
assumptions it can be claimed that the unidentified items can be ignored
for diet studies as the weight proportions of the individual food items
would be the same with and without the unidentified items. I am not
aware of published studies that quantify bias for the two approaches.

The allocation of unidentified items to identified items is not simple
in practise. A stomach may for example contain remains of an
unidentified fish and identified invertebrates. In such case it would
not be sensible to allocate the remains to the identified preys. However
other stomachs from the same predator and size class from the same haul
might include identified fish preys which could be used a "distribution
key" for unidentified items. If a given haul does not include identified
preys for a "distribution key" an alternative key could be obtained from
the neighbouring hauls or from another size classes within the haul.

It is not possible to give specific rules on how to allocate
unidentified prey items, other than the distribution key should be
derived from "local" data as far as possible. By "local" is meant from
the same haul, time period, predator size class, ICES rectangle,
roundfish area *etc*.

#### Assigning prey group ####
Stomach contents data may include prey identified to species species level, even for preys that 
are considered as "other prey" in the later data set produces for multi species assessments. 

With 250.000 stomachs sampled in the North Sea as basis for multispecies
models a pragmatic approach was needed for allocation unidentified
items. The former used NODC codes for preys are not maintained anymore
by NOAA, however NODC codes are still very useful for compilation of
stomach data as they are based on a 10-digit \"intelligent\" code. By
\"intelligent\" code numbers it is meant that information about taxonomy
was built into the codes through the use of 2-digit couplets to
represent one or more levels of the taxonomic hierarchy. For example, a
species assigned a 10-digit code would belong to the genus represented
by the first 8 digits of the code. In the compilation of stomach data,
this hierarchy makes it easier to assign (by an algorithm) partly
identified food items, to the identified items within the same genus,
family or order. Lookup table exist in the ICES system to assign a NODC
code to the WoRMS AphiaID code system presently used by ICES. The
section "Compilation of stomach data for use in the North Sea SMS model"
provides examples on how allocation of unidentified items could be done.

To be more specific. For the North sea data compilation the model prey species (e.g. cod, whiting, sprat)
were maintained as named preys. The remaining prey items were grouped according to the table below. Invertebrates (NODC < 8500000000)
are treated as "other food" in the model an aggregated into one group. Two groups of prey items, Clupeidae and Gadoidae were maintained as they were used to name e.g. preys that 
most likely were sprat or herring (Clupeidae) but could not be determined to species level. "other fish" includes fish preys that are not either a Clupeidae or Gadoidae. The "unknown" group (NODC > 900000000) was used for unidentified prey remains (either fish or invertebrates).  

 Species group	First NODC	Last NODC
 -------------  ----------  ---------
 other food      400000000	8499999999
 other fish	    8500000000	8747009999
 Clupeidae	    8747010000	8747019999
 other fish	    8747020000	8791029999
 Gadidae	      8791030000	8791031999
 Other fish	    8791032000	8999999999
 unknown	      9000000000	9999999999

to allocate the partly identified preys, preys within the Clupeidae and Gadidae were first allocated to the identified species within the group (if such exists)

### 7) Aggregation of stomachs contents into population metrics ###

The average diet should basically be calculated as a stratified mean of
the individual stomach content samples, weighted by strata abundances
(indices) of the predator and strata areas. For the initial compilation
of the "Year of the stomach" an average stomach contents was first
calculate by ICES rectangles and finally aggregated to population level
(ICES, 1997). When there was more than one sample within a rectangle the
rectangle value was calculated as the weighted mean of the sample (haul)
average values, the weighting factors being the number of stomachs from
each sample. Four methods were tried to calculate the "population"
average stomach:

i.  The rectangle values were averaged disregarding the number of
    stomachs sampled within each rectangle.

ii. The rectangle values were weighted by the sample size (total number
    of stomachs sampled within the rectangle).

iii. The rectangle values were weighted by the survey catch rates within
    the rectangle.

iv. The rectangle values were weighted by the square root of survey
    catch rates within the rectangle; this reduces the influence of
    occasional very large catch rates

The compilation (ICES, 1997) ended up using option *iv* for the
predators cod, haddock, whiting and mackerel, while option *ii* was used
for saithe because the samples (survey catches) were patchily
distributed.

The compilation of data for the SMS
model deviates from the original method used. Average stomach contents
by ICES rectangle are derived in the same way as done initially, but the SMS
method calculates then the weighted mean of stomach contents by ICES
roundfish area using the square root of the mean rectangle survey cpue
(option iv) as weighting factor (but option ii are used for saithe).

The calculation of the population diet for SMS was done as a weighted average
of the strata (roundfish area) average diet weighted by the product of
the average strata survey cpue and area of the strata.



References
==========

Andersen N.G., Beyer J.E. 2005a. Mechanistic modelling of gastric
evacuation applying the square root model to describe surface-dependent
evacuation in predatory gadoids. J Fish Biol 67:1392--1412.

Andersen N.G., Beyer J.E. 2005b. Gastric evacuation of mixed stomach
contents in predatory gadoids -- an expanded application of the square
root model to estimate food rations. Journal of Fish Biology
67:1413--1433.

Berg, C. W., A. Nielson, and K. Kristensen. 2014. Evaluation of
alternative age-based methods for estimating relative abundance from
survey data in relation to assessment models. Fisheries Research
151:91--99.

Daan, N., Bromley, P.J., Hislop, J.R.G and Nielsen, N.A. 1990. Ecology
of North Sea fish. Netherlands Journal of Sea Research,
26(2--4):343--386.

ICES. 1989. Database report of the Stomach Sampling Project 1981. Coop.
Res. Rep. 164: 1--145.

ICES. 1997. Database report of the Stomach Sampling Project 1991. Coop.
Res. Rep. 219: 1--442.

ICES. 1997. Report of the Multispecies Assessment Working Group, ICES
headquarters, 11--19 August 1997. ICES CM 1997/Assess:16, 235 pp.

Pinnegar, J.K. (2014) DAPSTOM - An Integrated Database & Portal for Fish
Stomach Records. Version 4.7. Centre for Environment, Fisheries &
Aquaculture Science, Lowestoft, UK. February 2014, 39 pp.



**\
**

**Appendix 1. ICES data exchange format for stomach data**
(<http://ices.dk/marine-data/data-portals/Pages/Fish-stomach.aspx> )

+-----------------------------------+-----------------------------------+
| **Field**                         | **Description**                   |
+-----------------------------------+-----------------------------------+
| Dataset                           | Dataset name                      |
+-----------------------------------+-----------------------------------+
| RecordType                        | SS for single stomach             |
+-----------------------------------+-----------------------------------+
| Country                           | Country that collected the data   |
+-----------------------------------+-----------------------------------+
| Ship                              | Vessel that collected the data    |
+-----------------------------------+-----------------------------------+
| Latitude                          | Data sampling position --         |
|                                   | latitude                          |
+-----------------------------------+-----------------------------------+
| Longitude                         | Data sampling position --         |
|                                   | longitude                         |
+-----------------------------------+-----------------------------------+
| Estimated\_Lat\_Long              | Flag whether the sampling         |
|                                   | position based on the reported    |
|                                   | area                              |
+-----------------------------------+-----------------------------------+
| ICES\_StatRec                     | ICES statistical rectangle        |
+-----------------------------------+-----------------------------------+
| ICES\_AreaCode                    | ICES area code                    |
+-----------------------------------+-----------------------------------+
| Year                              | YYYY                              |
+-----------------------------------+-----------------------------------+
| Month                             | MM                                |
+-----------------------------------+-----------------------------------+
| Day                               | DD                                |
+-----------------------------------+-----------------------------------+
| Time                              | Sampling time like HHMM           |
+-----------------------------------+-----------------------------------+
| Station                           | Station reference                 |
+-----------------------------------+-----------------------------------+
| Haul                              | Haul number                       |
+-----------------------------------+-----------------------------------+
| Sampling\_Method                  | Predator sampling method          |
+-----------------------------------+-----------------------------------+
| Depth                             | Sampling depth                    |
+-----------------------------------+-----------------------------------+
| Temperature                       | ^o^ C                             |
+-----------------------------------+-----------------------------------+
| SampleNo(FishID)                  | Predator reference code           |
|                                   |                                   |
|                                   | Fish ID unique for Country, year, |
|                                   | quarter and ship                  |
+-----------------------------------+-----------------------------------+
| ICES\_SampleID                    | ICES predator reference           |
+-----------------------------------+-----------------------------------+
| Predator\_AphiaID                 | Predator WoRMS AphiaID            |
+-----------------------------------+-----------------------------------+
| Predator\_LatinName               | Predator taxon Latin Name         |
+-----------------------------------+-----------------------------------+
| Predator\_Weight(mean)            | (Mean) predator weight            |
+-----------------------------------+-----------------------------------+
| Predator\_Age(mean)               | (Mean) predator age               |
+-----------------------------------+-----------------------------------+
| Predator\_Lengh(mean)             | (Mean) predator length            |
+-----------------------------------+-----------------------------------+
| Predator\_LowerLengthBound        | Predator's length lower bound     |
+-----------------------------------+-----------------------------------+
| Predator\_UpperLengthBound        | Predator's length upper bound     |
+-----------------------------------+-----------------------------------+
| Predator\_CPUE                    | Predator catch per hour           |
+-----------------------------------+-----------------------------------+
| GallBladder\_stage(class)         | Gall bladder stage                |
+-----------------------------------+-----------------------------------+
| Stomach\_METFP                    | Method of stomach preservation    |
+-----------------------------------+-----------------------------------+
| Stomach\_TotalNo                  | Total number of stomachs in the   |
|                                   | pool -- for single stomachs       |
|                                   | always 1.                         |
+-----------------------------------+-----------------------------------+
| Stomach\_WithFood                 | Number of stomachs with food      |
+-----------------------------------+-----------------------------------+
| Stomach\_Regurgitated             | Number of stomachs regurgitated   |
+-----------------------------------+-----------------------------------+
| Stomach\_WithSkeletalRemains      | Number of stomachs with skeletal  |
|                                   | remains                           |
+-----------------------------------+-----------------------------------+
| Stomach\_Empty                    | Number of empty stomachs          |
+-----------------------------------+-----------------------------------+
| Stomach\_ContentWgt               | Stomach content weight            |
+-----------------------------------+-----------------------------------+
| Stomach\_EmptyWgt                 | Stomach empty weight              |
+-----------------------------------+-----------------------------------+
| Stomach fullness                  | Stomach fullness                  |
+-----------------------------------+-----------------------------------+
| Stomach\_Item                     | Stomach item name (see Appendix   |
|                                   | B)                                |
+-----------------------------------+-----------------------------------+
| ICES\_ItemID                      | ICES stomach item ID              |
+-----------------------------------+-----------------------------------+
| Prey\_AphiaID                     | Prey WoRMS AphiaID                |
+-----------------------------------+-----------------------------------+
| Prey\_LatinName                   | Prey taxon Latin Name             |
+-----------------------------------+-----------------------------------+
| Prey\_IdentMet                    | Prey identification method        |
+-----------------------------------+-----------------------------------+
| Prey\_DigestionStage              | Prey digestion stage              |
|                                   |                                   |
|                                   | 0= Intact prey (skin, fins, legs  |
|                                   | and flesh is complete), 1=        |
|                                   | partially digested prey (prey in  |
|                                   | more advanced stages of           |
|                                   | digestion), 2= partially digested |
|                                   | prey (prey in more advanced       |
|                                   | stages of digestion), 3= skeletal |
|                                   | material (no flesh, only bones,   |
|                                   | shells, otoliths)                 |
+-----------------------------------+-----------------------------------+
| Prey\_TotalNo                     | Total number of preys             |
+-----------------------------------+-----------------------------------+
| Prey\_Weight                      | Prey weight in grams              |
+-----------------------------------+-----------------------------------+
| Prey\_LengthIdentifier            | Prey length identifier            |
+-----------------------------------+-----------------------------------+
| Prey\_Length                      | Prey length in cm                 |
+-----------------------------------+-----------------------------------+
| Prey\_LowerLengthBound            | Prey length lower bound           |
+-----------------------------------+-----------------------------------+
| Prey\_UpperLengthBound            | Prey length upper bound           |
+-----------------------------------+-----------------------------------+
| Prey\_MinNo                       | Minimum number of preys           |
+-----------------------------------+-----------------------------------+
| Remarks                           | Any relevant comments             |
+-----------------------------------+-----------------------------------+

**\
**

**Appendix b. Data exchange format used by the North Sea stomach
sampling program 1981-1991,**

Source *ICES Coop. Res. Rep. No. 219 , pp 421-422*

  **Position**   **Name**                      **Type^a^**   **Range of values**   **Comment**
  -------------- ----------------------------- ------------- --------------------- --------------------------------------------
  1-2            Ecosystem name                2N            1-8                   Footnote^b^ for coding scheme
  3-4            Year                          2N            01-99                 Year - 1900
  5              Quarter                       1N            1-4                   X
  6-9            Square                        4AN                                 ICES rectangle
  10-19          Predator code                 10N                                 NODC 10-digit 
  20-24          Sample number                 5N            1-99999               Unique fish ID
  25-27          Country                       3A                                  ICES alpha codes, No data:XXX
  28-31          Ship                          4A                                  ICES alpha, if available, No data: XXXX
  32-34          Sampling method               3A                                  Footnote^d^ for coding scheme No data: XXX
  35-40          Station/haul                  6AN                                 Use national system,No data: XXXXXX
  41-42          Month                         2N            01-12                 Not known: 99
  43-44          Day                           2N            01-31                 Not known: 99
  45-48          Time of day                   4N            0-2359,               Local time, hh/mm, start of tow. Not known: 9999
  49             Quadrant                      lN            1-4, 9                See footnote^d^ Not known: 9
  50-53          Latitude                      4N            0-9000,               dd/mm. Not known: 9999
  54-58          Longitude                     5N            0-18000,              ddd/mm. Not known: 99999
  59-61          Depth (meters)                3N            1-999                 Mean depth of tow. Not know: 999
  62-64          Temperature (bottom)          3N            -9.9 to 99.8          XX.X one implied decimal. Not known: 999
  65-68          Predator (mean) length        4N            1-9999                mm XXXX
  69-73          Predator (mean) weight        5N            1-99999               grams XXXXX. Not known: 99999
  74-75          Predator (mean) age           2N            0-99                  Not known: 99
  76-79          Predator lower length bound   4N            1-9999                mm XXXX
  80-83          Predator upper length bound   4N            1-9999                mm XXXX
  84-90          CPUE                          7N            1-9999999             Weighting coefficient for sample.Not known: 1
  91-93          Number with food              3N            0-999                 0, 1 for individual samples
  94-96          Number regurgitated           3N            0-999                 0, 1 for individual samples
  97-99          Number with skeletal remains  3N            0-999                 0, 1 for individual samples
  100-102        Number empty                  3N            0-999                 0, 1 for individual samples
  103-112        Prey species code             10N                                 NODC 10-digits
  113-116        Prey lower length bound       4N            1-9999                mm XXXX. Not known: 9999
  117-120        Prey upper length bound       4N            1-9999                mm XXXX. Not known: 9999
  121-128        Prey weight                   8N            1-99999999            Total weight mg XXXXXXXX
  129-134        Prey number                   6N            1-999999              Total number. Not known: 999999
  135            Digestion stage               1N            0-2, 9                See footnote^e^

a)  All numeric fields (N) right justified, zero filled; all apha (A)
    and mixed alpha/numeric fields (AN) left justified, space filled

b)  North Sea=l, Baltic Sea=2, Barents Sea=3, Iceland=4, Northeastern
    Newfoundland=5, Northeastern USA=6, Southern Shelf=7, Faroes=8

c)  DEM=demersally caught by trawling or seining gears PEL=pelagically
    caught by trawling or seining gears DHL=demersal hook and line

    PHL=pelagic hook and line DGN=demersal gill net PGN=pelagic gill net

d)  Quadrants for position identification are: l=NE, 2=NW, 3=SW, 4=SE
    (The axes of the quadrants are the equator and the Greenwich
    meridian)

e)  Digestion stages are: 0 =Pristine, 1 =Affected by digestion,
    2=Skeletal remains, 9=Unknown
