---
title: "Example 04: Population diet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ex04_Population_diet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  options(rmarkdown.html_vignette.check_title = FALSE),
  fig.width=7, fig.height=8
)
```


The population diet (estimated in Example 03) is organised as an object of the class STOMdiet, which is basically a list of predator and prey related information. The attributes to a STOMdiet (the control object) includes information on the compilation of data. 

```{r ex04_01}
library(FishStomachs)
suppressMessages(library(dplyr))

load(file=file.path(system.file( package = "FishStomachs"),"extdata","ex03d.Rdata"))
d

d[['PRED']]  # predator related information
d[['PREY']]  # prey related information

# show parts of the control attributes
print(attr(d,'control'),show=c('calc_diet','stratification'))


```
STOMdiet objects can be sub-setted and combined by the methods subset and c. 

```{r ex04_02}

# make a subset of data
d1<-subset(d,pred_size=='0400-0499' & stratum_time=="1991-Q1")
d1

# another subset
d2<-subset(d, stratum_time=="1991-Q2")
d2

#combine subsets
c(d1,d2)

rm(d1,d2)
```

It is also possible to transform STOMdiet object into a plain data.frame, modify data and convert back again.

```{r ex04_03}

# convert into a data frame
df<-as.data.frame(d)
df

# add a variable 
df$cannibal<-as.character(df$pred_name)==as.character(df$prey_name)

#convert it back to a STOMdiet object with a new PREY variable
d2<-as_STOMdiet(df,new_prey_var = 'cannibal')

# "cannibal" s now included (in the PREY data set)
names(d2[['PREY']])

rm(d2)

```
There are various summary functions available for the STOMdiet object. 

```{r ex04_04}

summary(d)
summary(d,level=2)

# make a subset of data
d1<-subset(d,pred_size=='0300-0399' & stratum_time=="1991-Q1")
d1
summary(d1,level=3)
summary(d1,level=4,drop.unused.levels=TRUE)

```

And plots.
```{r ex04_05}

#default
plot(d)

# change in layout, shorter predator size on the x-axis ("from" value),
#   added number of stomachs and added title
plot(d,Ncol=1,cut_pred_size=c(1,4),addTitle = TRUE, addNstom=TRUE,refac_prey=TRUE)

```


It is possible to change the names of both predator and prey species.

```{r ex04_06}

# Change names into English names
d2<-from_to_species_diet(d,pred_from_to=c("species","short"),prey_from_to=c("species","short"),
          sp_info_file=file.path(system.file( package = "FishStomachs"),"extdata",'species_info.csv'))
                         
plot(d2,Ncol=1,cut_pred_size=c(1,4),addTitle = TRUE, addNstom=TRUE)




```
